---
title: "brainflowprobes users guide"
author: 
  - name: Amanda Price
    affiliation:
    - &libd Lieber Institute for Brain Development, Johns Hopkins Medical Campus
    email: amanda.joy.price@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('brainflowprobes')`"
vignette: >
  %\VignetteIndexEntry{brainflowprobes users guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()

## Bib setup
library('knitcitations')

## Load knitcitations with a clean bibliography
cleanbib()
cite_options(hyperlink = 'to.doc', citation_format = 'text', style = 'html')
# Note links won't show for now due to the following issue
# https://github.com/cboettig/knitcitations/issues/63

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation('BiocStyle'),
    derfinder = citation('derfinder')[1],
    derfinderPlot = citation('derfinderPlot')[1],
    sessioninfo = citation('sessioninfo'),
    GenomicRanges = citation('GenomicRanges'),
    knitcitations = citation('knitcitations'),
    knitr = citation('knitr')[3],
    rmarkdown = citation('rmarkdown'),
    rtracklayer = citation('rtracklayer'),
    testthat = citation('testthat'),
    ggplot2 = citation('ggplot2'),
    cowplot = citation('cowplot'),
    BSgenome.Hsapiens.UCSC.hg19 = citation('BSgenome.Hsapiens.UCSC.hg19'),
    Biostrings = citation('Biostrings'),
    RColorBrewer = citation('RColorBrewer'),
    bumphunter = citation('bumphunter')[1]
)

write.bibtex(bib, file = 'brainflowprobes_ref.bib')
```

# Basics

## Install `r Biocpkg('brainflowprobes')`

`R` is an open-source statistical environment which can be easily modified to enhance its functionality via packages. `r Biocpkg('brainflowprobes')` is a `R` package available via the [Bioconductor](http://bioconductor/packages/brainflowprobes) repository for packages. `R` can be installed on any operating system from [CRAN](https://cran.r-project.org/) after which you can install `r Biocpkg('brainflowprobes')` by using the following commands in your `R` session:

```{r 'installDer', eval = FALSE}
install.packages("BiocManager")
BiocManager::install("brainflowprobes")

## Check that you have a valid Bioconductor installation
BiocManager::valid()
```

## Required knowledge

`r Biocpkg('brainflowprobes')` is based on many other packages and in particular in those that have implemented the infrastructure needed for dealing with RNA-seq data. That is, packages like `r Biocpkg('GenomicFeatures')` and `r Biocpkg('rtracklayer')` that allow you to import the data. A `r Biocpkg('brainflowprobes')` user is not expected to deal with those packages directly but will need to be familiar with `r Biocpkg('SummarizedExperiment')` to understand the results `r Biocpkg('brainflowprobes')` generates. It might also prove to be highly beneficial to check the


* `r Biocpkg('derfinder')` package for performing annotation-agnostic differential expression analysis.

If you are asking yourself the question "Where do I start using Bioconductor?" you might be interested in [this blog post](http://lcolladotor.github.io/2014/10/16/startBioC/#.VkOKbq6rRuU).

## Asking for help

As package developers, we try to explain clearly how to use our packages and in which order to use the functions. But `R` and `Bioconductor` have a steep learning curve so it is critical to learn where to ask for help. The blog post quoted above mentions some but we would like to highlight the [Bioconductor support site](https://support.bioconductor.org/) as the main resource for getting help: remember to use the `recount` tag and check [the older posts](https://support.bioconductor.org/t/brainflowprobes/). Other alternatives are available such as creating GitHub issues and tweeting. However, please note that if you want to receive help you should adhere to the [posting guidelines](http://www.bioconductor.org/help/support/posting-guide/). It is particularly critical that you provide a small reproducible example and your session information so package developers can track down the source of the error.

## Citing `r Biocpkg('brainflowprobes')`

We hope that `r Biocpkg('brainflowprobes')` will be useful for your research. Please use the following information to cite the package and the overall approach. Thank you!

```{r 'citation'}
## Citation info
citation('brainflowprobes')
```

# Quick start to using to `r Biocpkg('brainflowprobes')`

TODO


# Reproducibility

The `r Biocpkg('brainflowprobes')` package `r citep(bib[['brainflowprobes']])` was made possible thanks to:

* R `r citep(bib[['R']])`
* `r Biocpkg('BiocStyle')` `r citep(bib[['BiocStyle']])`
* `r Biocpkg('Biostrings')` `r citep(bib[['Biostrings']])`
* `r Biocpkg('BSgenome.Hsapiens.UCSC.hg19')` `r citep(bib[['BSgenome.Hsapiens.UCSC.hg19']])`
* `r Biocpkg('bumphunter')` `r citep(bib[['bumphunter']])`
* `r CRANpkg('cowplot')` `r citep(bib[['cowplot']])`
* `r Biocpkg('derfinder')` `r citep(bib[['derfinder']])`
* `r Biocpkg('derfinderPlot')` `r citep(bib[['derfinderPlot']])`
* `r CRANpkg('sessioninfo')` `r citep(bib[['sessioninfo']])`
* `r Biocpkg('GenomicRanges')` `r citep(bib[['GenomicRanges']])`
* `r CRANpkg('ggplot2')` `r citep(bib[['ggplot2']])`
* `r CRANpkg('knitcitations')` `r citep(bib[['knitcitations']])`
* `r CRANpkg('knitr')` `r citep(bib[['knitr']])`
* `r CRANpkg('RColorBrewer')` `r citep(bib[['RColorBrewer']])`
* `r Biocpkg('rtracklayer')` `r citep(bib[['rtracklayer']])`
* `r CRANpkg('rmarkdown')` `r citep(bib[['rmarkdown']])`
* `r CRANpkg('testthat')` `r citep(bib[['testthat']])`
    
    
Code for creating the vignette

```{r createVignette, eval=FALSE}
## Create the vignette
library('rmarkdown')
system.time(render('brainflowprobes-vignette.Rmd', 'BiocStyle::html_document'))

## Extract the R code
library('knitr')
knit('brainflowprobes-vignette.Rmd', tangle = TRUE)
```

```{r createVignette2}
## Clean up
file.remove('brainflowprobes_ref.bib')
```

Date the vignette was generated.

```{r reproduce1, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r reproduce2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits=3)
```

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library('sessioninfo')
options(width = 120)
session_info()
```



# Bibliography

This vignette was generated using `r Biocpkg('BiocStyle')` `r citep(bib[['BiocStyle']])`
with `r CRANpkg('knitr')` `r citep(bib[['knitr']])` and `r CRANpkg('rmarkdown')` `r citep(bib[['rmarkdown']])` running behind the scenes.

Citations made with `r CRANpkg('knitcitations')` `r citep(bib[['knitcitations']])`.

```{r vignetteBiblio, results = 'asis', echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
bibliography()
```
